---
title: "`tlverse`: Implement frameworks, not algorithms"
date: "`r format(Sys.time(), '%Y %b %d (%a), %H:%M')`"
output:
  xaringan::moon_reader:
    lib_dir: libs
    chakra: libs/remark-latest.min.js
    css: ["default", "custom.css"]
    nature:
      highlightStyle: zenburn
      highlightLines: true
      navigation: {
        scroll: false
      }
    self-contained: true
---

```{r knitr_setup, echo=FALSE, message=FALSE}
knitr::opts_chunk$set(fig.width = 7, fig.height = 4.5, dpi = 300,
                      fig.cap = "", fig.align = "center")
showtext::showtext.opts(dpi = 300)
library(sl3)
library(methods)
```

class: center, middle

# `tlverse`: The Targeted Learning Analytics Ecosystem

---

# The `tlverse` Ecosystem

What is the `tlverse`?

By analogy to the [`tidyverse`](https://tidyverse.org/):

> The tidyverse is an opinionated collection of R packages designed for data
> science. All packages share an underlying design philosophy, grammar, and data
> structures. 

So, the [`tlverse`](https://tlverse.org) is

* an opinionated collection of R packages for Targeted Learning
* sharing an underlying philosophy, grammar, and set of data structures

---

# The `tlverse` Ecosystem

These are the main packages that represent the **core** of the `tlverse`:

--

* [`sl3`](https://github.com/tlverse/sl3): Modern Super Learning with Pipelines
  * _What?_ A modern object-oriented re-implementation of the Super Learner
    algorithm, employing recently developed paradigms for `R` programming.
  * _Why?_ A design that leverages modern tools for fast computation, is
    forward-looking, and can form one of the cornerstones of the `tlverse`.

--

* [`tmle3`](https://github.com/tlverse/tmle3): An Engine for Targeted Learning
  * _What?_ A generalized framework that simplifies Targeted Learning by
    identifying and implementing a series of common statistical estimation
    procedures.
  * _Why?_ A common interface and engine that accommodates current algorithmic
    approaches to Targeted Learning and is still flexible enough to remain the
    engine even as new techniques are developed.

---

# The `tlverse` Ecosystem

In addition to the engines that drive development in the `tlverse`, there are
some supporting packages -- in particular, we have two...

--

* [`origami`](https://github.com/tlverse/origami): A Generalized Framework for
   Cross-Validation
  * _What?_ A generalized framework for flexible cross-validation
  * _Why?_ Cross-validation is a key part of ensuring error estimates are honest
    and preventing overfitting. It is an essential part of the both the Super
    Learner algorithm and Targeted Learning.

--

* [`delayed`](https://github.com/tlverse/delayed): Parallelization Framework for
   Dependent Tasks
  * _What?_ A framework for delayed computations (futures) based on task
    dependencies.
  * _Why?_ Efficient allocation of compute resources is essential when deploying
    large-scale, computationally intensive algorithms.


---

# The Problem

--

Mark is too productive. Even for point treatment data, we have a huge range of
TMLE methodologies:

* **TMLE Variants:** TMLE, IPCW-TMLE, CV-TMLE, C-TMLE, One-step TMLE, HAL-TMLE,
   Targeted HAL, etc.

* **Parameters:** Means under static interventions, dynamic rules, stochastic
  interventions, ATE, ATT, CDE, NDE, Blip Variance, etc.

--

For each TMLE-based method, someone has probably implemented the estimator at
least well enough to run simulations for a paper. However, those implementations
are not often maintained after the paper is published, and they might not have
been terribly robust in the first place (e.g., `opttx`). Thus, these methods are
not easy for the average user to apply.

---

# The Solution: `tlverse` + `tmle3`

* Identify the parts of the framework common across a range of methods, and
   implement those.

* Build tools to let others build on the framework.

* Build a unified and accessible user interface to these methods.

* Establish a good software development community so that the project can
   persist past the interest of any one developer.

---
# It's Roadmap Time!

<u>**Steps 1-3: Define the research question**</u>

**Data:** $O = (W, A, Y) \sim P_0 \in \mathcal{M}$, where, for simplicity,
consider $W \in \mathbb{R}^d$, $A \in \{0,1\}$, $Y\in \{0,1\}$.

**Model:** $\mathcal{M}$, a nonparametric model that makes no assumptions on the
distribution governing the data $P_0$. 

We consider the following _nonparametric structural equation model_ (NPSEM): 

$$W = f_W(U_W)$$
$$A = f_A(W, U_A)$$
$$Y = f_Y(W, U_Y)$$

Which corresponds to the likelihood factorization
$P_0 = P_0(Y \mid A,W)P_0(A \mid W)P_0(W)$

**Parameter:** Treatment Specific Mean $\Psi(P) = E_W[E[Y \mid A = 1, W]]$

---
# It's Roadmap Time!

<u>**Steps 4-6: Estimation**</u>

**Super Learner (Likelihood Estimation):** Identify and estimate the relevant
likelihood factors with Super Learner

**TMLE:** Update our initial likelihood estimates to solve the efficient
influence function (EIF), apply the parameter mapping to the updated likelihood.

In addition to the elements defined in steps 1-3, this requires the following:

* **EIF** -- for TSM $\frac{I(A = 1)}{P(A = 1 \mid W)}(Y - E[Y \mid A, W]) + E[Y \mid A = 1 ,W] - \psi$
* **submodel** -- we'll use a logistic submodel
* **loss function** -- we'll use log-likelihood loss
* **solver** (iterative vs. one-step) -- not relevant for the TSM parameter

**Inference:** Use the variance of the EIF to construct confidence intervals:
$$\sigma^2 = \frac{1}{n}\sum_{i = 1}^n D^2(O_i) + o_p\left(\frac{1}{\sqrt{n}}\right)$$

---

# Example: Define the Research Question

We use data from the Collaborative Perinatal Project (CPP), available in the
`sl3` package. To simplify this example, we define a binary intervention
variable, `parity01` -- an indicator of having one or more children before the
current child and a binary outcome, `haz01` -- an indicator of having an above
average height for age.

```{r setup, echo=FALSE, message=FALSE}
library(sl3)
library(tmle3)
library(data.table)
```

```{r define data}
data(cpp_imputed)
data <- as.data.table(cpp_imputed)

# generate binary treatment and outcome
data$parity01 <- as.numeric(data$parity > 0)
data$haz01 <- as.numeric(data$haz > 0)
data[is.na(data)] <- 0

# define variable roles
node_list <- list(
  W = c("apgar1", "apgar5", "gagebrth", "mage",
                "meducyrs", "sexn"),
  A = "parity01",
  Y = "haz01"
)
```

---

# Example: Super Learning with `sl3`

```{r cpp_sl}

# define regression tasks
Q_task <- make_sl3_Task(data, 
                        covariates=c(node_list$W, node_list$A), 
                        outcome=node_list$Y)
g_task <- make_sl3_Task(data, 
                        covariates=node_list$W, 
                        outcome=node_list$A)

# simple sl
stack <- make_learner_stack(Lrnr_glm, Lrnr_mean)
lrnr_nnls <- make_learner(Lrnr_nnls)
lrnr_sl <- make_learner(Lrnr_sl, stack, lrnr_nnls)

Q_fit <- lrnr_sl$train(Q_task)
g_fit <- lrnr_sl$train(g_task)

```

---

# A Simple TMLE Implementation

```{r tmle_tsm}

# get relevant quantities
A <- data$parity01
Y <- data$haz01

#P(A=1|W)
g1W <- g_fit$predict(g_task)

#E(Y|A=a,W)
QAW <- Q_fit$predict(Q_task)

#create counterfactual data and make task
cf_data <- copy(data)
set(cf_data, , node_list$A, 1)
cf_Q_task <- make_sl3_Task(cf_data, 
                           covariates=c(node_list$W, node_list$A), 
                           outcome=node_list$Y)

#E(Y|A=1,W)
Q1W <- Q_fit$predict(cf_Q_task)
```

---

# A Simple TMLE Implementation

```{r tmle_tsm_2}

####
# construct clever covariate
# I(A=1|W)
HA = A/g1W

# 1/P(A=1|W)
H1 = 1/g1W

####
# fit logistic submodel
submodel_fit <- glm(Y ~ HA - 1 + offset(qlogis(QAW)), family=binomial())
epsilon <- coef(submodel_fit)

####
# update likelihood 
Q1W_star <- plogis(qlogis(Q1W) + H1*epsilon)
QAW_star <- plogis(qlogis(QAW) + HA*epsilon)
```

---

# A Simple TMLE Implementation

```{r tmle_tsm_3}
####
# calculate IC
IC <- HA * (Y - QAW_star) + Q1W_star - mean(Q1W_star)

####
# verify convergence
mean(IC)

####
# get estimate
psi_hat <- mean(Q1W_star)
print(psi_hat)
```

---

# Summary

* We implemented a simple TMLE in ~ 50 lines of code.

* But we haven't done much to implement TMLE _in general_

* Lots of things are hardcoded, or not explictly coded at all, including the
   NPSEM, the submodel, the parameter, and the loss function.

--

* `tmle3` takes a different approach to this problem

* Defining a TMLE method in `tmle3` loosely follows the roadmap (order is a bit
   different)

* Most parts of it are _modular_ in that they can be easily replaced to
   implement slightly different TMLEs

---

# Model: `tmle3_Node`s

In `tmle3`, we define the NPSEM using the `define_node` function for each node.
`define_node` allows a user to specify the node_name, which columns in the data
comprise the node, and a list of parent nodes. 
``` {r define_npsem}
npsem <- list(
  define_node("W", c(
    "apgar1", "apgar5", "gagebrth", "mage",
    "meducyrs", "sexn"
  )),
  define_node("A", c("parity01"), c("W")),
  define_node("Y", c("haz01"), c("A", "W"))
)
```

Nodes also track information about the data types of the variables (continuous,
categorical, binomial, etc). Here, that information is being estimated
automatically from the data. In the future, each node will also contain
information about censoring indicators, where applicable, but this is not yet
implemented.

---

# Data: `tmle3_task`

A `tmle3_Task` is an object comprised of observed data, and the NPSEM defined
above:

```{r tmle3_data}
tmle_task <- tmle3_Task$new(data, npsem = npsem)
```

This task object contains methods to help subset the data as needed for various
steps in the tmle process:

```{r subset_task}
#get the outcome node data
head(tmle_task$get_tmle_node("Y"))
```

A `tmle3_Task` is a special kind of `sl3` task.

---

# Likelihood Fits: `Likelihood`

`tmle3` models likelihoods as a list of likelihood factor objects, where each
likelihood factor object describes either _a priori_ knowledge or an estimation
strategy for the corresponding likelihood factor. These objects all inherit from
the `LF_base` base class, and there are different types depending on which of a
range of estimation strategies or _a priori_ knowledge is appropriate.

In some cases, a full conditional density for a particular factor is not
necessary. Instead, a conditional mean (a much easier quantity to estimate), is
all that's required. Although conditional means are not truly likelihood
factors, conditional means are also modeled using using likelihood factor
objects.

Examples:

* `LF_fit` - estimate a likelihood factor using `sl3` learners
* `LF_emp` - estimate a likelihood factor using NP-MLE (marginal likelihood
   only)

---

# Likelihood Fits: `Likelihood`

Going back to our CPP example, we will estimate the marginal likelihood of $W$,
using NP-MLE, the conditional density of $A$ given $W$ using a GLM fit via `sl3`
and the conditional mean of $Y$ given $A$ and $W$ using another GLM fit via
`sl3`:

```{r likelihood}
# define and fit likelihood
factor_list <- list(
  define_lf(LF_emp, "W"),
  define_lf(LF_fit, "A", lrnr_sl),
  define_lf(LF_fit, "Y", lrnr_sl, type="mean")
)

likelihood_def <- Likelihood$new(factor_list)
likelihood <- likelihood_def$train(tmle_task)
print(likelihood)
```

A `tmle3` `Likelihood` is actually a special type of `sl3` learner, so the
syntax to train it on data is analogous.

---

# Likelihood Fits: `Likelihood`

Having fit the likelihood, we can now get likelihood values for any
`tmle3_task`:

```{r likelihood values}
likelihood_values <- likelihood$get_likelihoods(tmle_task)
head(likelihood_values)
```

---

# Counterfactuals: `CF_Likelihood`


In `tmle3`, interventions are modeled by likelihoods where one or more
likelihood factors is replaced with a counterfactual version representing some
intervention. 

`tmle3` defines the `CF_Likelihood` class, which inherits from `Likelihood`, and
takes an `observed_likelihood` and an `intervention_list`. 

For our CPP example, we'll define a simple intervention where we set all
treatment $A=1$. We do this by defining a static likelihood factor, as a simple
indicator function $P(A \mid W) = I(A = 1)$

```{r intervention}
intervention <- define_lf(LF_static, "A", value = 1)
```

We can then use this to construct a counterfactual likelihood:

```{r cf_likelihood}
cf_likelihood <- make_CF_Likelihood(likelihood, intervention)
```

A `cf_likelihood` is a likelihood object, and so has the same behavior as the
observed likelihood object defined above, but with the observed likelihood
factors being replaced by the defined intervention likelihood factors.

---

# Counterfactuals: `CF_Likelihood`

In particular, we can get likelihood values under the counterfactual likelihood:

```{r cf_likelihood values}
cf_likelihood_values <- cf_likelihood$get_likelihoods(tmle_task)
head(cf_likelihood_values)
```

We see that the likelihood values for the $A$ node are all either 0 or 1, as
would be expected from an indicator likelihood function. In addition, the
likelihood values for the non-intervention nodes have not changed.
---

# Parameter: `Param_base`

In the TMLE framework, we define a target parameter $\Psi(P)$ as a mapping from
a probability distribution $P \in \mathcal{M}$ to a set of real numbers
$\mathbb{R}^d$. Here $\mathcal{M}$ is implied by the NPSEM we defined above. 

In `tmle3`, we define parameter objects as objects inheriting from the
`Param_base` class, which keep track of not only the mapping from a probability
distribution to a parameter value, but also the corresponding EIF of the
parameter, and the "clever covariates" needed to calculate a TMLE update to the
likelihood. These values are calculated using the `tmle3_task`, `Likelihood`,
and `CF_likelihood` objects defined above.


Here, we define a treatment specific mean (TSM) parameter based on the
intervention we defined previously:

```{r param}
tsm <- define_param(Param_TSM, likelihood, intervention)
```

This structure will be documented so that new parameters can be implemented
easily, analogous to learners in `sl3`.

---

# Update

The update procedure component of `tmle3` is currently in flux. The current
structure is as follows: We have an object, `tmle3_Update`, which calculates the
individual update steps using `tmle3_Update$update_step`. This contains the
submodel and loss function needed to solve the EIF. 

```{r update}
updater <- tmle3_Update$new(tsm)
likelihood$update_list <- updater
```

This is currently the weak link in `tmle3`, and it's under active development to
make it faster and more elegant.

---

# Fit

Now that we have specified all the components required for the TMLE procedure,
we can generate an object that manages all the components and updates the
likelihood and solves the EIF.

```{r fit_tmle}
tmle_fit <- fit_tmle3(tmle_task, likelihood, tsm, updater)
print(tmle_fit)
```

Currently, the solver is iterative, but we plan to extend it to accomodate the
one-step approach.

---

# User-Friendly Interface

* The `tmle3` framework described above is general, and allows most components of
   the TMLE procedure to be specified in a modular way.

* However, most end-users will not be interested in manually specifying all of
   these components.

* Therefore, `tmle3` implements a `tmle3_Spec` object that bundles a set of
   components into a _specification_ that, with minimal additional detail, can
   be run by an end-user:

---

# User-Friendly Interface

```{r tmle_spec}
nodes <- list(W = c("apgar1", "apgar5", "gagebrth", "mage",
                  "meducyrs", "sexn"),
              A = "parity01",
              Y = "haz01")

learner_list <- list(Y = lrnr_sl, A = lrnr_sl)


data2 <- data.table::copy(data) # make a new copy to deal with data.table

tmle_fit_from_spec <- tmle3(tmle_TSM_all(), data2, nodes, learner_list)
print(tmle_fit_from_spec)
```

---

# Summary

* We have tried to explicitly model all the steps of the roadmap, including all
   the different components necessary to specify and implement TMLE methods.

* We believe this results in a general framework that can implement most, if not
   all, of the TMLE methods being developed.

* Much more work needs to be done on both the framework and building out the
   various TMLE methods on top of the framework.

* Success will depend on engagement with the broader Targeted Learning community
   -- We'd appreciate any feedback or contributions you can offer!
